Fractional Decomposition Tree Algorithm: A tool for studying the integrality gap of Integer Programs

### 2022-01-17 FDT2020/main-section.tex
first run through reviewer 1 comments

Consider an IP instance $I=(A,b)$. Define sets $S(I)$ and $P(I)$ as in (\ref{S}) and (\ref{P}), respectively. Assume $S(I)\subseteq \{0,1\}^n$ and $P(I)\subseteq [0,1]^n$. For simplicity in the notation we assume an instance $I$ and denote $P(I),S(I),$ and $g(I)$ by $P$, $S$, and $g$ for this section and the next section. Also, for both sections let $x^*$ be the optimal solution to the LP formulation and assume $t=|\spp(x^*)|$. Reorder variables as necessary to put all the nonzero variables into the first $t$ indices. So we can assume $x^*_i = 0$ for $i=t+1,\ldots,n$.
%\cindy{TODO: Is this something that should be clear to the reader with what they know at this point? Does it need an explanation? Does it need an explanation at the end of this section or the next?}\arash{this is just a relabeling of the indices.} \cindy{Sorry.  I didn't understand that you meant the solution has that property. Right? The vector $x$ isn't defined in this paragraph and it's not in the next lemma either. If $x$ is the output of DomToIP, then we should say that.  ``Without loss of generality the solution $x$ returned by DomToIP has ...''}\arash{I see your point, please see if the change made it better} \cindy{Yes, thanks.  We can delete this conversation after you see this response.}

In this section we prove Theorem \ref{domtoIP}. In fact, we prove a stronger result. 
\begin{lemma}\label{domlemma}
	Given an integral vector in the dominant of an LP relaxation ($\tilde{x}\in \dom(P)$ and $\tilde{x}\in \{0,1\}^n$) for an IP instance that has integrality gap $g < \infty$, there is an algorithm (the DomToIP algorithm) that finds $\bar{x}\in S$ in polynomial time such that $\bar{x}\leq \tilde{x}$.\end{lemma}
Lemma \ref{domlemma} implies Theorem \ref{domtoIP}, since it is easy to obtain an integer point in $\dom(P)$: numerically rounding up $x^*$ (or any fractional point in $P$) gives us a point in $\dom(P)$.
%\cindy{I added the caveat that the instance has a finite integrality gap and made it a little more self-contained.  Is it OK?}
Hence, we can assume in the proof below that $\tilde{x}_i= 0$ for $i=t+1,\ldots,n$. 
%\arash{I also made a change based on the change above}\cindy{Thanks!}
\subsection{Proof of Lemma \ref{domlemma}: The DomToIP Algorithm}

@@

We prove that indeed $x^{(t)}\in S$. First, we need to show that in any iteration $\ell=  0,\ldots,t-1$ of DomToIP that $\HelperLP(x^{(\ell)})$ is feasible. We show something stronger. For $\ell=0,\ldots,t-1$ let
\begin{align*}
\LP^{(\ell)}&= \{z\in P\; : \; z\leq x^{(\ell)} \mbox{ and } z_j=x_j^{(\ell)} \mbox{ for } j\in [\ell]\}, \text{ and}\\
\IP^{(\ell)}&= \{z\in \LP^{(\ell)}\; : \; z\in \{0,1\}^n\}.
\end{align*}
If $\LP^{(\ell)}$ is a non-empty set then $\HelperLP(x^{(\ell)})$ is feasible. We show by induction on $\ell$ that $\LP^{(\ell)}$ and $\IP^{(\ell)}$ are not empty sets for $\ell=0,\ldots,t-1$. First, $\LP^{(0)}$ is feasible since by definition $x^{(0)}\in \dom(P)$, meaning there exists $z\in P$ such that $z\leq x^{(0)}$. By Theorem \ref{CV2}, there exists $\tilde{z}^i\in S$ and $\theta_i\geq 0$ for $i\in [k]$ such that $\sum_{i=1}^{k} \theta_i = 1$ and $\sum_{i=1}^{k}\theta_i \tilde{z}^i \leq gz$. (This assumes a finite integrality gap $g$).
%\cindy{Is this the place (and others like it) where we require $g$ to be finite?}
Hence, $\sum_{i=1}^{k}\theta_i \tilde{z}^i \leq gz\leq gx^{(0)}$. So if $x^{(0)}_j=0$, then $ \sum_{i=1}^{k}\theta_i \tilde{z}_j^i =0$, which implies that $\tilde{z}^i_j=0$ for all $i\in [k]$ and $j\in [n]$ where $x^{(0)}_j=0$. Hence, $\tilde{z}^i\leq x^{(0)}$ for $i\in [k]$. Therefore $\tilde{z}^i\in \IP^{(0)}$ for $i\in [k]$, which implies $\IP^{(0)}\neq \emptyset$.

@@

We consider two cases. In the first case, we have $z^*_{\ell+1}=0$. In this case we set $x^{(\ell+1)}_{\ell+1}=0$. Since $z^*\leq x^{(\ell+1)}$, we have $z^*\in \LP^{(\ell+1)}$. Also, $z^*\in P$. By Theorem \ref{CV2} there exists $\tilde{z}^i\in S$ and $\theta_i\geq 0$ for $i\in [k]$ such that $\sum_{i=1}^{k} \theta_i = 1$ and  $\sum_{i=1}^{k}\theta_i \tilde{z}^i \leq gz^*$. We have $\sum_{i=1}^{k}\theta_i \tilde{z}^i \leq gz^*\leq gx^{(\ell+1)}$.
So for $j\in [n]$ where $x^{(\ell+1)}_j=0$, we have $z^i_j=0$ for $i\in [k]$. This implies $\tilde{z}^i\leq x^{(\ell+1)}$ for $i=1,\ldots,k$. Hence, there exists $z\in S$ such that $z\leq x^{(\ell+1)}$. We claim that $z\in \IP^{(\ell+1)}$. If $z\notin \IP^{(\ell+1)}$ we must have $1\leq j \leq \ell$ such that $z_j < x^{(\ell+1)}_{j}$, and thus $z_j = 0$ and $x^{(\ell+1)}_j=1$. Without loss of generality assume $j$ is the minimum index (between $1$ and $\ell$) satisfying $z_j < x^{(\ell+1)}_{j}$. Consider iteration $j$ of the DomToIP algorithm. We have $z\leq x^{(\ell+1)}\leq x^{(j)}$.
%\cindy{I certainly believe you and Bob on correctness, and I was convinced of the correctness of (and wrote a bunch of the proofs for) the earlier version.  But while I have a little time, I'll point out places where I was confused (and hence a reader might be). This last statement I don't see right away.  It doesn't seem that $x^{(\ell+1)}\leq x^{(j)}$ in general when $j < \ell + 1$. It seems that in general, for a fractional element $x_q$, DomToIP can round the $q$th element either up or down depending on whether it's OK to round down (future feasibility assured).  So it seems that $x^j$, which is more fractional, could have some elements smaller and some elements larger than the more integral $x^{(\ell+1)}$.  This confusion carries forward to the assertion that $z$ is a feasible solution to $\HelperLP(x^{(j-1)})$.  Suppose that $x_q = .2$ for some $q < \ell$ and Dom2IP rounded the $q$th element up to $1$. $\HelperLP(x^{q})$ has a constraint $z_q \le 0.2$, but subsequent LPs, such as $\HelperLP(x^{(\ell)})$ have a constraint $x_q = 1$. So the later LP solutions may not be feasible for the earlier ones.}
We also have $x^{(j)}_j=1$ which implies when we solved $\HelperLP(x^{(j-1)})$ the optimal value was strictly larger than zero. However, $z$ is a feasible solution to $\HelperLP(x^{(j-1)})$ and gives an objective value of 0. This is a contradiction, so $z\in \IP^{(\ell+1)}$.

@@

	Let $y^0$, $y^1$, $\lambda^*_0$, and $\lambda^*_1$ be an optimal solution to the LP above. Set $\gamma_0=\lambda^*_0$ and $\gamma_1 = \lambda^*_1$.
%Let $x^0$, $x^1$, $\lambda_0$, and $\lambda_1$ be an optimal solution to the LP above. Set $\gamma_0=\lambda_0$ and $\gamma_1 = \lambda_1$.
\cindy{The sentence right before this is how I think you link the $\lambda$ in the LP to the $\gamma$ constants in the Lemma statement and the claim below, etc.}
Let $\hat{x}^0 = y^0/\lambda^*_0$, $\hat{x}^1=y^1/\lambda^*_1$. This choice satisfies  (ii), (iii), (iv). To show that (i) is also satisfied we prove the following claim.
	\arash{Now that I notice, we use $\lambda_0,\lambda_1,x^0,x^1$ to represnt a feasible solution to the system 13-18. Since the optimal solution is use only for the duration of one sentence above maybe we can change the notation for the optimal solution. For instance use $y^0,y^1, \lambda^*_0,\lambda^*_1$}
\cindy{I think I made the changes you suggested.  I also used the LP optima to set the $\hat{x}$.  Please check.}
	\begin{claim}\label{CVexists}
		We have $\gamma_0 + \gamma_1\geq 1/g$.
	\end{claim}
	\begin{cproof}
		We show that there is a feasible solution that achieves the objective value of $\frac{1}{g}$. By Theorem \ref{CV2} there exists $\theta \in [0,1]^k$, with $\sum_{i=1}^{k}\theta_i = 1$ and $\tilde{x}^i\in S$ for $i\in[k]$ such that 
		$\sum_{i=1}^{k}\theta_i \tilde{x}^i\leq gx'$. So

>>>>

Consider an IP instance $I=(A,b)$. Define sets $S(I)$ and $P(I)$ as in (\ref{S}) and (\ref{P}), respectively. Assume $S(I)\subseteq \{0,1\}^n$ and $P(I)\subseteq [0,1]^n$. For simplicity in the notation we assume an instance $I$ and denote $P(I),S(I),$ and $g(I)$ by $P$, $S$, and $g$ for this section and the next section. Also, for both sections let $x^*$ be the optimal solution to the LP formulation and assume $t=|\spp(x^*)|$. Reorder variables as necessary to put all the nonzero variables into the first $t$ indices. So we can assume $x^*_i = 0$ for $i=t+1,\ldots,n$.

Before diving into the main results, we state the following simple observation that we will use several times in the proofs of this section and the next section.
\begin{observation}\label{dom01}
Any point $x\in [0,1]^n$ that is a convex combination of points in $\dom(S)$, is also a convex combination of points in $\dom(S) \cap [0,1]^n$
\end{observation}
\begin{proof}
This is because if there exists ${z}^i\in \dom(S)$ and $\theta_i\geq 0$ for $i\in [k]$ such that $\sum_{i=1}^{k} \theta_i = 1$ and $\sum_{i=1}^{k}\theta_i {z}^i \leq gz$, then 
we let ${y}^i$ be the coordiante-wise minimum of vector ${z}^i$ and $\{1\}^n$ for $i\in [k]$ and we have ${y}^i\in \dom(S)$ and $\sum_{i=1}^{k}\theta_i{y}^i \leq gz$.
\end{proof}
%\cindy{TODO: Is this something that should be clear to the reader with what they know at this point? Does it need an explanation? Does it need an explanation at the end of this section or the next?}\arash{this is just a relabeling of the indices.} \cindy{Sorry.  I didn't understand that you meant the solution has that property. Right? The vector $x$ isn't defined in this paragraph and it's not in the next lemma either. If $x$ is the output of DomToIP, then we should say that.  ``Without loss of generality the solution $x$ returned by DomToIP has ...''}\arash{I see your point, please see if the change made it better} \cindy{Yes, thanks.  We can delete this conversation after you see this response.}

In this section we prove Theorem \ref{domtoIP}. In fact, we prove a stronger result. 
\begin{lemma}\label{domlemma}
	Given an integral vector $\tilde{x}$ in the dominant of an LP relaxation ($\tilde{x}\in \dom(P)$ and $\tilde{x}\in \{0,1\}^n$) for an IP instance that has integrality gap $g < \infty$, there is an algorithm (the DomToIP algorithm) that finds $\bar{x}\in S$ in polynomial time such that $\bar{x}\leq \tilde{x}$.\end{lemma}
Lemma \ref{domlemma} implies Theorem \ref{domtoIP}, since it is easy to obtain an integer point in $\dom(P)$: numerically rounding up $x^*$ (or any fractional point in $P$) gives us a point in $\dom(P)$.
%\cindy{I added the caveat that the instance has a finite integrality gap and made it a little more self-contained.  Is it OK?}
Hence, we can assume in the proof below that $\tilde{x}_i= 0$ for $i=t+1,\ldots,n$. 
%\arash{I also made a change based on the change above}\cindy{Thanks!}
\subsection{Proof of Lemma \ref{domlemma}: The DomToIP Algorithm}

@@

We prove that indeed $x^{(t)}\in S$. First, we need to show that in any iteration $\ell=  0,\ldots,t-1$ of DomToIP that $\HelperLP(x^{(\ell)})$ is feasible. We show something stronger. For $\ell=0,\ldots,t-1$ let
\begin{align*}
\LP^{(\ell)}&= \{z\in P\; : \; z\leq x^{(\ell)} \mbox{ and } z_j=x_j^{(\ell)} \mbox{ for } j\in [\ell]\}, \text{ and}\\
\IP^{(\ell)}&= \{z\in \LP^{(\ell)}\; : \; z\in \{0,1\}^n\}.
\end{align*}

If $\LP^{(\ell)}$ is a non-empty set then $\HelperLP(x^{(\ell)})$ is feasible. We show by induction on $\ell$ that $\LP^{(\ell)}$ and $\IP^{(\ell)}$ are not empty sets for $\ell=0,\ldots,t-1$.

First, $\LP^{(0)}$ is feasible since by definition $x^{(0)}\in \dom(P)$, meaning there exists $z\in P$ such that $z\leq x^{(0)}$. By Theorem \ref{CV2} and Observation \ref{dom01}, there exists $\tilde{z}^i\leq 1$ in $\dom(S)$ and $\theta_i\geq 0$ for $i\in [k]$ such that $\sum_{i=1}^{k} \theta_i = 1$ and $\sum_{i=1}^{k}\theta_i \tilde{z}^i \leq gz$. (This assumes a finite integrality gap $g$).
%\cindy{Is this the place (and others like it) where we require $g$ to be finite?}

 Hence, $\sum_{i=1}^{k}\theta_i \tilde{z}^i \leq gz\leq gx^{(0)}$. So if $x^{(0)}_j=0$, then $ \sum_{i=1}^{k}\theta_i \tilde{z}_j^i =0$, which implies that $\tilde{z}^i_j=0$ for all $i\in [k]$ and $j\in [n]$ where $x^{(0)}_j=0$. Hence, $\tilde{z}^i\leq x^{(0)}$ for $i\in [k]$. Therefore $\tilde{z}^i\in \IP^{(0)}$ for $i\in [k]$, which implies $\IP^{(0)}\neq \emptyset$.

@@

We consider two cases. In the first case, we have $z^*_{\ell+1}=0$. In this case we set $x^{(\ell+1)}_{\ell+1}=0$. Since $z^*\leq x^{(\ell+1)}$, we have $z^*\in \LP^{(\ell+1)}$. Also, $z^*\in P$. By Theorem \ref{CV2} and Observation \ref{dom01}, there exists $\tilde{z}^i\leq 1$ in $\dom(S)$ and $\theta_i\geq 0$ for $i\in [k]$ such that $\sum_{i=1}^{k} \theta_i = 1$ and  $\sum_{i=1}^{k}\theta_i \tilde{z}^i \leq gz^*$. We have $\sum_{i=1}^{k}\theta_i \tilde{z}^i \leq gz^*\leq gx^{(\ell+1)}$.
So for $j\in [n]$ where $x^{(\ell+1)}_j=0$, we have $z^i_j=0$ for $i\in [k]$. This implies $\tilde{z}^i\leq x^{(\ell+1)}$ for $i=1,\ldots,k$. Hence, there exists $z\in S$ such that $z\leq x^{(\ell+1)}$. We claim that $z\in \IP^{(\ell+1)}$. If $z\notin \IP^{(\ell+1)}$ we must have $1\leq j \leq \ell$ such that $z_j < x^{(\ell+1)}_{j}$, and thus $z_j = 0$ and $x^{(\ell+1)}_j=1$. Without loss of generality assume $j$ is the minimum index (between $1$ and $\ell$) satisfying $z_j < x^{(\ell+1)}_{j}$. Consider iteration $j$ of the DomToIP algorithm. We have $z\leq x^{(\ell+1)}\leq x^{(j)}$.
%\cindy{I certainly believe you and Bob on correctness, and I was convinced of the correctness of (and wrote a bunch of the proofs for) the earlier version.  But while I have a little time, I'll point out places where I was confused (and hence a reader might be). This last statement I don't see right away.  It doesn't seem that $x^{(\ell+1)}\leq x^{(j)}$ in general when $j < \ell + 1$. It seems that in general, for a fractional element $x_q$, DomToIP can round the $q$th element either up or down depending on whether it's OK to round down (future feasibility assured).  So it seems that $x^j$, which is more fractional, could have some elements smaller and some elements larger than the more integral $x^{(\ell+1)}$.  This confusion carries forward to the assertion that $z$ is a feasible solution to $\HelperLP(x^{(j-1)})$.  Suppose that $x_q = .2$ for some $q < \ell$ and Dom2IP rounded the $q$th element up to $1$. $\HelperLP(x^{q})$ has a constraint $z_q \le 0.2$, but subsequent LPs, such as $\HelperLP(x^{(\ell)})$ have a constraint $x_q = 1$. So the later LP solutions may not be feasible for the earlier ones.}
We also have $x^{(j)}_j=1$ which implies when we solved $\HelperLP(x^{(j-1)})$ the optimal value was strictly larger than zero. However, $z$ is a feasible solution to $\HelperLP(x^{(j-1)})$ and gives an objective value of 0. This is a contradiction, so $z\in \IP^{(\ell+1)}$.

@@

	Let $y^0$, $y^1$, $\lambda^*_0$, and $\lambda^*_1$ be an optimal solution to the LP above. Set $\gamma_0=\lambda^*_0$ and $\gamma_1 = \lambda^*_1$.
%Let $x^0$, $x^1$, $\lambda_0$, and $\lambda_1$ be an optimal solution to the LP above. Set $\gamma_0=\lambda_0$ and $\gamma_1 = \lambda_1$.
\cindy{The sentence right before this is how I think you link the $\lambda$ in the LP to the $\gamma$ constants in the Lemma statement and the claim below, etc.}
Let $\hat{x}^0 = y^0/\lambda^*_0$, $\hat{x}^1=y^1/\lambda^*_1$. This choice satisfies  (ii), (iii), (iv). To show that (i) is also satisfied we prove the following claim.
	\arash{Now that I notice, we use $\lambda_0,\lambda_1,x^0,x^1$ to represnt a feasible solution to the system 13-18. Since the optimal solution is use only for the duration of one sentence above maybe we can change the notation for the optimal solution. For instance use $y^0,y^1, \lambda^*_0,\lambda^*_1$}
\cindy{I think I made the changes you suggested.  I also used the LP optima to set the $\hat{x}$.  Please check.}
	\begin{claim}\label{CVexists}
		We have $\gamma_0 + \gamma_1\geq 1/g$.
	\end{claim}
	\begin{cproof}
		We show that there is a feasible solution that achieves the objective value of $\frac{1}{g}$. By Theorem \ref{CV2} and Observation \ref{dom01} there exists $\theta \in [0,1]^k$, with $\sum_{i=1}^{k}\theta_i = 1$ and $\tilde{x}^i\leq 1$ in $\dom(S)$ for $i\in[k]$ such that 
		$\sum_{i=1}^{k}\theta_i \tilde{x}^i\leq gx'$. So
